Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-14 ThesisJP/Background/background.tex
writing and added figure.

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI-express bus, which connects the different parts of a computer to the motherboard. This is the reason why the first rules of GPGPU is to minimize PCI-express traffic.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

@@

 the most powerful one (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).

@@

@@

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies above 2 MHz are typically used.

@@

\begin{figure}[t!]
\subfigure[Delay-and-sum]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}
}
\subfigure[Capon]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}
}
\caption{Array beam pattern with uniform and Capon weights. An interfering source is located at 20 degrees.}
\end{figure}

>>>>

For the first adopters of GPU programming, or General-purpose programming on graphics processing units (GPGPU) as it was named, it soon became evident that offloading all types of computations to the GPU was not always a good idea. In most cases it is important to balance the load equally between the CPU and GPU to obtain their combined computationally power. Balancing typically means that highly parallel tasks should be offloaded to the GPU and serialized work should stay on the CPU. This paradigm of utilizing specialized cores for solving specific problems is today known as heterogeneous computing. The specialized core can be a GPU or e.g. a field-programmable gate array (\nom{FPGA}{Field-programmable gate array}). A recent trend is that specialized accelerator cores are embedded onto the CPU. This heterogeneous processors is often referred to as an accelerated processing unit (\nom{APU}{Accelerated processing unit}). The last three generations of CPUs from Intel have all had on-chip integrated graphics. Even if an on-chip GPU is less powerful than a discrete GPU, valuable time is saved by not having to send data across the PCI Express bus, which connects the different parts of a computer to the motherboard. This is the reason why the first rules of GPGPU (with discrete GPUs) is to minimize CPU-to-GPU transfers.

The final transition currently taking place in the computer industry is the introduction of cloud computing, where compute power is provided as a web-service and automatically scales on demand. Together, multi-core, heterogeneous cores, and cloud cores forms the "jungle" that today's software engineers have to tackle (Fig.\,\ref{fig:jungle}). This thesis focus on heterogeneous computing with one multi-core CPU and a discrete high performance GPU, a cheap and high performance system which is likely to be the specification of current and future ultrasound imaging system.

\subsection{Comparing CPU and GPU performance}\label{sec:cpu_vs_gpu}
The CPU is design to be the main control unit of a computer. It is therefore good at general processing. The GPU, on the other hand, has a history of processing graphics which involves heavy vector arithmetic. Even though they started out as very different processors, modern CPUs are including more and more GPU-like features, and GPUs are adding CPU features. In this section we will see why there is a big difference in peak performance for a CPU and GPU.

It is common to say that a CPU has few but advance generic cores, and the GPU has many but simple cores.

@@

 the most powerful integrated intel GPU (HD 5200) have a peak performance of nearly one tera single precision floating point operations per second (FLOPs).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{img/CPU-GPU.png}
\caption{Overview of CPU and GPU architecture.}
\label{fig:cpu_gpu}
\end{figure}

@@

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.

@@

\begin{figure}[t!]
\subfigure[Delay-and-sum]{
	\includegraphics[width=0.47\textwidth]{img/scenario_das_resp2.png}
}
\subfigure[Capon]{
	\includegraphics[width=0.47\textwidth]{img/scenario_mv_resp2.png}
}
\caption{Array beam pattern with uniform and Capon weights. An interfering source is located at 20\degree. Notice the reduced sidelobe level for the Capon beam pattern in the direction of the interferer.}
\end{figure}
